Parameter precision in economic growth models inspired by Ecological Economics

### 2015-10-09 Articles/Article_EcolEcon_2015/Paper0/Paper0-Resampling.Rnw
Edit text to eliminate sigma in favor of rho.

**** Go completely generic everywhere ($x_1$ instead of $k$). ****

The Cobb-Douglas production function 
is the canonical representation of indexed economic output ($y$) 
as a function of indexed factors of production~($x_1$, $x_2$, and $x_3$) and 
an additional exogenous, time-dependent term~($A$).
This production function can be expressed as
%
\begin{equation} \label{eq:CD}
  y = \theta A \, {x_1}^{\alpha_1} \, {x_2}^{\alpha_2} \, {x_3}^{\alpha_3} \; ; \; A \equiv \mathrm{e}^{\lambda t},
\end{equation}
%
where 
$y \equiv Y/Y_{0}$,
$\theta$ is a scale parameter,
e is the base of the natural logarithm, 
$\lambda$ is represents the pace of technological progress,
$t$ is time measured in years,
$x_1 = k \equiv K/K_{0}$, 
$x_2 = l \equiv L/L_{0}$, 
$x_3 = e \equiv E/E_{0}$, 
$Y$ is economic output represented by GDP, 
$K$ is capital stock expressed in currency units, 
$L$ is labor expressed in workers or work-hours/year, 
$E$ is energy expressed in units of energy per time, typically TJ/year, and
0 subscripts indicate values at an initial year.
All historical data~($y$, $k$, $l$, and $e$) are indexed (by ratio)
relative to an initial year. 
Time~($t$) is indexed (by difference) relative to an initial year.
After indexing, $\theta$ is expected to be near unity,
but no other parameters change.%
  \footnote{Dimensionless, indexed quantities are represented by lower-case symbols 
  ($y$, $x$, $k$, $l$, and $e$), 
  and dimensional quantities are represented by upper-case symbols 
  ($Y$, $K$, $L$, and $E$). 
  Model parameters are represented by Greek letters
  ($\theta$, $\lambda$, $\alpha_1$, $\alpha_2$, $\alpha_3$, and
  later $\delta_1$, $\delta$, $\sigma_1$, and $\sigma$).}

@@

The Constant Elasticity of Substitution~(CES) model 
is given as
%
\begin{equation} \label{eq:CESgeneric}
  y = \theta \: A \: \left\{\delta B_1^{\rho/\rho_1} 
      + (1-\delta) x_3^{-\rho} \right\}^{-1/\rho}; 
      A \equiv \mathrm{e}^{\lambda t},
\end{equation}
%
where
$B_1 \equiv \delta_1 x_1^{-\rho_1} + (1-\delta_1) x_2^{-\rho_1}$, and
$x_1$, $x_2$, and $x_3$ are factors of production
and permutations of capital~($k$), labor~($l$), and energy~($e$).
The $(kl)e$ nesting has $x_1 = k$, $x_2 = l$, and $x_3 = e$; 
the $(le)k$ nesting has $x_1 = l$, $x_2 = e$, and $x_3 = k$; and
the $(ek)l$ nesting has $x_1 = e$, $x_2 = k$, and $x_3 = l$.

@@

To estimate the parameters $\theta$, $\lambda$, $\alpha_1$, $\alpha_2$, and $\alpha_3$ 
in the Cobb-Douglas model, 
we reparameterized Equation~\ref{eq:CD} as
%
\begin{equation} \label{eq:CDe_AB_reparam}
  y = \theta \, \mathrm{e}^{\lambda t} \, x_1^{\alpha_1} \, x_2^{\alpha_2} \, x_3^{1 - \alpha_1 - \alpha_2}
\end{equation}
%
ensuring that $\alpha_1 + \alpha_2 + \alpha_3 = 1$,
thereby providing constant returns to scale.
The \texttt{R} \citep{R} function \texttt{lm} was used to estimate 
values of $\lambda$, $\theta$, $\alpha_1$, and $\alpha_2$ in log-transform space.
If the fitted value for $\alpha_1$ or $\alpha_2$ fell outside the interval $[0,1]$,
we fit along all boundaries 
($\alpha_1 = 1$,
$\alpha_2 = 1$,
$\alpha_3 = 1$,
$\alpha_1 = 0$,
$\alpha_2 = 0$, and
$\alpha_3 = 0$)
and chose the boundary fit with
minimum sum of squared errors~(\sse) as the winning model:
%
\begin{equation} \label{eq:sse}
  \sse = \sum_{i} r_i^2 \; ,
\end{equation}
%
where the residual is given as
%
\begin{equation} \label{eq:residual}
  r_i \equiv \ln (y_i / \hat{y}_i)
\end{equation}
%
is the $i^{\mathrm{th}}$ residual (all models assume multiplicative errors),
and
$\hat{y}_i$ is the fitted value for economic output at time $t_i$.
$\alpha_3$ was recovered with $\alpha_3 = 1 - \alpha_1 - \alpha_2$.

In cases where a model adheres to the cost share principle,
output elasticities are fixed
in accordance with long-run historical cost shares
for capital and labor ($\hat{\alpha}_1 = 0.3$ and $\hat{\alpha}_2 = 0.7$).
Under these conditions, $\hat{\alpha}_3$ is assumed to be $0$.

@@

Next, a gradient search (using \texttt{PORT} and \texttt{L-BFGS-B} algorithms available in \texttt{micEconCES})
is attempted wherein all parameters
($\theta$, $\lambda$, $\delta_1$, $\delta$, $\rho_1$, and $\rho$)
are allowed to float.
The starting values for fitting parameters are taken from
the grid search point that provided the lowest sum of squared errors~(\sse).

\citet{Henningsen:2011td},
in their detailed analysis of Kemfert~\cite{kemfert1998estimated}, found that
%
\begin{quote}
\ldots the Levenberg-Marquardt and the \texttt{PORT} algorithms
are---at least in this study---most likely to find the
coefficients that give the best fit to the model,
where the \texttt{PORT} algorithm can be used to restrict the estimates
to the economically meaningful region.
\end{quote}
%
In our testing, we found that to be mostly true.
\texttt{PORT} nearly always provides lower
\sse{} than \texttt{L-BFGS-B}, despite the fact that
\texttt{L-BFGS-B} often reports convergence
and \texttt{PORT} does not for the same data
(i.e., for the same $y$, $k$, $l$, and $e$ time series).

In addition to the above trials,
we also fit along all boundaries of the economically-meaningful region
of the CES model.
Boundaries are given by all combinations of the following:
$\delta_1$~=~0~or~1,
$\delta$~=~0~or~1,
$\sigma_1$~=~0~or~$\infty$, and
$\sigma$~=~0~or~$\infty$.
Table~\ref{tab:CES_boundary_models} shows the set
of 20 degenerate equations found along parameter boundaries~(rows~1--20),
wherein factors of production ($x_1$,~$x_2$,~$x_3$)
are permutations of ($k$,~$l$,~$e$).
The bottom (unnumbered) row of Table~\ref{tab:CES_boundary_models}
shows the generic three-factor CES equation.
Each degenerate equation provides a boundary model.
Note that many boundary parameter combinations
yield equivalent CES boundary models,%
  \footnote{
  For example, all boundary parameter combinations
  that include $\delta = 0$ yield the same boundary model, $y = \theta A x_3$.
  }
so there are far fewer boundary models~(20) than the maximum possible number
of degenerate equations~(81).%
  \footnote{
  There are three possible states (unspecified, lower boundary, upper boundary)
  for four constrained parameters ($\delta_1$, $\sigma_1$, $\delta$, $\sigma$),
  which gives $3^4 = 81$ possible boundary models.
  }
In the parameter columns of Table~\ref{tab:CES_boundary_models},
``---'' indicates that a parameter is unknowable on that boundary
(because it does not appear in the boundary model and
it is not part of the definition of the boundary), and
an empty space indicates that the parameter
is estimable on that boundary using the degenerate equation.

Fitting along boundaries is important for two reasons.
First, bounded, non-linear, ordinary least squares~(OLS) algorithms
that perform gradient searches within the parameter space
(such as the \texttt{PORT} and \texttt{L-BFGS-B} algorithms discussed above)
often have difficulties dealing with boundaries.
Fitting directly \emph{on} the boundaries
of the economically-meaningful region
ensures that a sum of squared errors~(\sse)
minimum located at a boundary will be found, if it exists.
Second, fitting on the boundaries of the economically-meaningful region
prevents erroneous reporting of unknowable parameters.
Table~\ref{tab:CES_boundary_models} includes many dashes~(``---'')
indicating that many fitting parameters are unknowable
in the boundary models.
For example, $\sigma_1$ and $\sigma$ are unknowable in
the boundary model shown in Row 1 of Table~\ref{tab:CES_boundary_models}.
If an OLS search algorithm
(such as \texttt{PORT} and \texttt{L-BFGS-B}, discussed above)
operating with the full CES model (bottom row of Table~\ref{tab:CES_boundary_models})
were to find an \sse{} minimum along that boundary,
it would report values for $\sigma_1$ and $\sigma$
\emph{in addition to} $\theta$, $\lambda$, $\delta_1$~(which will be unity),
and $\delta$~(also unity).
Under these conditions,
it is clearly erroneous to report
meaningless, because they are unknowable,
values for $\sigma_1$ and $\sigma$.
Fitting with the boundary models
shown in Table~\ref{tab:CES_boundary_models}
avoids this mistake.

@@

When $\sigma_1$ or $\sigma$ is 0,

@@

  E.g., when $\sigma_1 = 0$ and $\delta~=~1$, the Leontief model is

@@

When $\sigma_1$ or $\sigma$ is $\infty$,

@@

  E.g., when $\sigma_1 = \infty$ and $\delta~=~1$, the linear model is

@@

      & $\delta_1$  & $\sigma_1$ & $\delta$   & $\sigma$ & Boundary model \\

@@

      4. & ---         & 0          & 1          & ---      & $y = \theta A \min

@@

      5. & 1           & ---        & ---        & 0        & $y = \theta A \min

@@

      6. & 0           & ---        & ---        & 0        & $y = \theta A \min

@@

      7. &             & $\infty$   & 1          & ---

@@

      8. & 1           & ---        &            & $\infty$

@@

      9. & 0           & ---        &            & $\infty$

@@

     13. & ---         & 0          & ---        & 0        & $y = \theta A \min

@@

     14. & ---         & 0          &            & $\infty$

@@

     15. &             & $\infty$   & ---        & 0

@@

     16. &             & $\infty$   &            & $\infty$

@@

     17. & ---         & 0          &            &

@@

     18. &            &            & ---     & 0

@@

     19. &             &            &            & $\infty$

@@

     20. &             & $\infty$   &            &

@@

\begin{equation} \label{eq:resampled_y}
  \tilde Y_i' = \hat y_i \pm r_j \; ,
\end{equation}
%
where
%
\authNote{Do we need this equation for $r_i$ here? We already have Equation~\ref{eq:residual}.}%
%
\begin{equation} \label{eq:residual_2}
  r_i = y_i - \hat y_i \; .
\end{equation}
%
Both the sign~($\pm$) and the index
($j$,~typically different from $i$)
of the residual~($r_j$)
are chosen at random (with replacement).
We repeated the resampling process 1000 times.

The parameters from the fit to a resampled time series
(the ``resample parameters'') will be different from the parameters
obtained from the fit to historical data (the ``base parameters'').
When these resample parameters are highly variable,
it is an indication
that the data do not determine the parameter estimates very precisely.
Lack of precision can stem from a number of factors, most obviously
a poor model fit,
low model sensitivity to one or more parameters, or
correlations between parameter estimates.

@@

We choose the resampling approach instead of the more-common technique
of estimating confidence intervals from standard errors for two reasons.
First, model parameters are highly constrained, and confidence intervals often violate the constraints.
For example, a result of $\alpha_1 = 0.25 \pm 0.3$ is nonsensical,
because $ \alpha_1 \in [0, 1]$.
Second, true confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha_1 = 0.2 + 0.1, -0.05$),
but the standard error approach yields symmetric confidence intervals.
The resampling approach that we employed both respects constraints
and allows for asymmetric confidence intervals.

@@

Refer to other paper for historica data.

>>>>

The Cobb-Douglas production function 
is the canonical representation of indexed economic output ($y$) 
as a function of indexed factors of production~($x_1$, $x_2$, and $x_3$) and 
an additional exogenous, time-dependent term~($A$).
This production function can be expressed as
%
\begin{equation} \label{eq:CD}
  y = \theta A \, {x_1}^{\alpha_1} \, {x_2}^{\alpha_2} \, {x_3}^{\alpha_3} \; ; \; A \equiv \mathrm{e}^{\lambda t},
\end{equation}
%
where 
$y \equiv Y/Y_{0}$,
$\theta$ is a scale parameter,
e is the base of the natural logarithm, 
$\lambda$ is the so-called Solow residual,
$t$ is time measured in years,
$x_1 = k \equiv K/K_{0}$, 
$x_2 = l \equiv L/L_{0}$, 
$x_3 = e \equiv E/E_{0}$, 
$Y$ is economic output represented by GDP, 
$K$ is capital stock expressed in currency units, 
$L$ is labor expressed in workers or work-hours/year, 
$E$ is energy expressed in units of energy per time, typically TJ/year, and
0 subscripts indicate values at an initial year.
All historical data~($y$, $k$, $l$, and $e$) are indexed (by ratio)
relative to an initial year. 
Time~($t$) is indexed (by difference) relative to an initial year.
Due to indexing, $\theta$ is expected to be near unity,
but no other parameters are affected.%
  \footnote{Dimensionless, indexed quantities are represented by lower-case symbols 
  ($y$, $x$, $k$, $l$, and $e$), 
  and dimensional quantities are represented by upper-case symbols 
  ($Y$, $K$, $L$, and $E$). 
  Model parameters are represented by Greek letters
  ($\theta$, $\lambda$, $\alpha_1$, $\alpha_2$, $\alpha_3$, and
  later $\delta_1$, $\delta$, $\sigma_1$, and $\sigma$).}

@@

The Constant Elasticity of Substitution~(CES) production function 
is given as
%
\begin{equation} \label{eq:CESgeneric}
  y = \theta \: A \: \left\{\delta B_1^{\rho/\rho_1} 
      + (1-\delta) x_3^{-\rho} \right\}^{-1/\rho}; 
      A \equiv \mathrm{e}^{\lambda t},
\end{equation}
%
where
$B_1 \equiv \delta_1 x_1^{-\rho_1} + (1-\delta_1) x_2^{-\rho_1}$, and
$x_1$, $x_2$, and $x_3$ are factors of production
and permutations of capital~($k$), labor~($l$), and energy~($e$).
The $(kl)e$ nesting has $x_1 = k$, $x_2 = l$, and $x_3 = e$; 
the $(le)k$ nesting has $x_1 = l$, $x_2 = e$, and $x_3 = k$; and
the $(ek)l$ nesting has $x_1 = e$, $x_2 = k$, and $x_3 = l$.

@@

To estimate the parameters $\theta$, $\lambda$, $\alpha_1$, $\alpha_2$, and $\alpha_3$ 
in the Cobb-Douglas model, 
we reparameterized Equation~\ref{eq:CD} as
%
\begin{equation} \label{eq:CDe_AB_reparam}
  y = \theta \, \mathrm{e}^{\lambda t} \, x_1^{\alpha_1} \, x_2^{\alpha_2} \, x_3^{1 - \alpha_1 - \alpha_2}
\end{equation}
%
ensuring that $\alpha_1 + \alpha_2 + \alpha_3 = 1$,
thereby providing constant returns to scale.
$\alpha_3$ was recovered with $\alpha_3 = 1 - \alpha_1 - \alpha_2$.
The \texttt{R} \citep{R} function \texttt{lm} was used to estimate 
values of $\lambda$, $\theta$, $\alpha_1$, and $\alpha_2$ in log-transform space.
If the fitted value for $\alpha_1$ or $\alpha_2$ fell outside the interval $[0,1]$,
we fit along all boundaries 
($\alpha_1 = 1$,
$\alpha_2 = 1$,
$\alpha_3 = 1$,
$\alpha_1 = 0$,
$\alpha_2 = 0$, and
$\alpha_3 = 0$)
and chose the boundary fit with
minimum sum of squared errors~(\sse) as the winning model:
%
\begin{equation} \label{eq:sse}
  \sse = \sum_{i} r_i^2 \; ,
\end{equation}
%
where
%
\begin{equation} \label{eq:residual}
  r_i \equiv \ln (y_i / \hat{y}_i)
\end{equation}
%
is the $i^{\mathrm{th}}$ residual (all models assume multiplicative errors),
and
$\hat{y}_i$ is the fitted value for economic output at time $t_i$.

In cases where a model adheres to the cost share principle,
output elasticities are fixed
in accordance with long-run historical cost shares
for capital and labor ($\hat{\alpha}_1 = 0.3$ and $\hat{\alpha}_2 = 0.7$).
Under these conditions, $\hat{\alpha}_3$ is assumed to be $0$.%
  \authNote{Need to fix the discussion of $\alpha$s and the CSP.}

@@

Next, a gradient search (using \texttt{PORT} and \texttt{L-BFGS-B} algorithms available in \texttt{micEconCES})
is attempted wherein all parameters
($\theta$, $\lambda$, $\delta_1$, $\delta$, $\rho_1$, and $\rho$)
are allowed to float.
The starting values for parameters are taken from
the grid search point that provided the lowest sum of squared errors~(\sse).

\citet{Henningsen:2011td},
in their detailed analysis of Kemfert~\cite{kemfert1998estimated}, found that
%
\begin{quote}
  \ldots the Levenberg-Marquardt and the \texttt{PORT} algorithms
  are---at least in this study---most likely to find the
  coefficients that give the best fit to the model,
  where the \texttt{PORT} algorithm can be used to restrict the estimates
  to the economically meaningful region.
\end{quote}
%
In our testing, we found that to be mostly true.
\texttt{PORT} nearly always provides lower
\sse{} than \texttt{L-BFGS-B}, despite the fact that
\texttt{L-BFGS-B} often reports convergence
and \texttt{PORT} does not for the same historical data
(i.e., for the same $y$, $k$, $l$, and $e$ time series).

In addition to the above trials,
we also fit along all boundaries of the economically-meaningful region
of the CES model.
Boundaries are given by all combinations of the following:
$\delta_1$~=~0~or~1,
$\delta$~=~0~or~1,
$\sigma_1$~=~0~or~$\infty$, and
$\sigma$~=~0~or~$\infty$.
Table~\ref{tab:CES_boundary_models} shows the set
of 20 degenerate equations found along parameter boundaries~(rows~1--20),
wherein factors of production ($x_1$,~$x_2$,~$x_3$)
are permutations of ($k$,~$l$,~$e$).
The bottom (unnumbered) row of Table~\ref{tab:CES_boundary_models}
is a repeat of Equation~\ref{eq:CESgeneric}.
Each degenerate equation provides a boundary model.
Note that many boundary parameter combinations
yield equivalent CES boundary models,%
  \footnote{
  For example, all boundary parameter combinations
  that include $\delta = 0$ yield the same boundary model, $y = \theta A x_3$.
  }
so there are far fewer boundary models~(20) than the maximum possible number
of degenerate equations~(81).%
  \footnote{
  There are three possible states (unspecified, lower boundary, upper boundary)
  for four constrained parameters ($\delta_1$, $\sigma_1$, $\delta$, $\sigma$),
  which gives $3^4 = 81$ possible boundary models.
  }
In the parameter columns of Table~\ref{tab:CES_boundary_models},
``---'' indicates that a parameter is unknowable on that boundary
(because it does not appear in the boundary model and
it is not part of the definition of the boundary), and
an empty space indicates that the parameter
is estimable on that boundary using the degenerate equation.

Fitting along boundaries is important for two reasons.
First, bounded, non-linear, ordinary least squares~(OLS) algorithms
that perform gradient searches within the parameter space
(such as the \texttt{PORT} and \texttt{L-BFGS-B} algorithms discussed above)
often have difficulties dealing with boundaries.
Fitting directly \emph{on} the boundaries
of the economically-meaningful region
ensures that a sum of squared errors~(\sse)
minimum located at a boundary will be found, if it exists.
Second, fitting on the boundaries of the economically-meaningful region
prevents erroneous reporting of unknowable parameters.
Table~\ref{tab:CES_boundary_models} includes many dashes~(``---'')
indicating that many parameters are unknowable
in the boundary models.
For example, $\sigma_1$ and $\sigma$ are unknowable in
the boundary model shown in Row 1 of Table~\ref{tab:CES_boundary_models}.
If an OLS search algorithm
(such as \texttt{PORT} and \texttt{L-BFGS-B}, discussed above)
operating with the full CES model (bottom row of Table~\ref{tab:CES_boundary_models})
were to find an \sse{} minimum along that boundary,
it would report values for $\rho_1$ and $\rho$
\emph{in addition to} $\theta$, $\lambda$, $\delta_1$~(which will be unity),
and $\delta$~(also unity).
Under these conditions,
it is clearly erroneous to report
meaningless, because they are unknowable,
values for $\rho_1$ and $\rho$.
Fitting with the boundary models
shown in Table~\ref{tab:CES_boundary_models}
avoids this mistake.

**** Change to talk about $\rho$ only. ****

@@

When $\rho_1$ or $\rho$ is $\infty$,

@@

  E.g., when $\rho_1 = \infty$ and $\delta~=~1$, the Leontief model is

@@

When $\rho_1$ or $\rho$ is $-1$,

@@

  E.g., when $\rho_1 = -1$ and $\delta~=~1$, the linear model is

@@

      & $\delta_1$     & $\rho_1$   & $\delta$   & $\rho$   & Boundary model \\

@@

      4. & ---         & $\infty$   & 1          & ---      & $y = \theta A \min

@@

      5. & 1           & ---        & ---        & $\infty$ & $y = \theta A \min

@@

      6. & 0           & ---        & ---        & $\infty$ & $y = \theta A \min

@@

      7. &             & $-1$       & 1          & ---

@@

      8. & 1           & ---        &            & $-1$

@@

      9. & 0           & ---        &            & $-1$

@@

     13. & ---         & $\infty$   & ---        & $\infty$   & $y = \theta A \min

@@

     14. & ---         & $\infty$   &            & $-1$

@@

     15. &             & $-1$       & ---        & $\infty$

@@

     16. &             & $-1$       &            & $-1$

@@

     17. & ---         & $\infty$   &            &

@@

     18. &            &            & ---         & $\infty$

@@

     19. &             &            &            & $-1$

@@

     20. &             & $-1$       &            &

@@

\begin{equation} \label{eq:resampled_y}
  \tilde Y_i' = \hat y_i \pm r_j \; ,
\end{equation}
%
where
%
\authNote{Do we need this equation for $r_i$ here? We already have Equation~\ref{eq:residual}.}%
%
\begin{equation} \label{eq:residual_2}
  r_i = y_i - \hat y_i \; .
\end{equation}
%
Both the sign~($\pm$) and the index
($j$,~typically different from $i$)
of the residual~($r_j$) in Equation~\ref{eq:resampled_y}
are chosen at random (with replacement).
We repeated the resampling process 1000 times.

The parameters from the fit to a resampled time series
(the ``resample parameters'') will be different from the parameters
obtained from the fit to historical data (the ``base parameters'').
When the distribution of these resample parameters 
(the ``bootstrap resample distribution'') 
is highly variable,
it is an indication
that the data do not determine the parameter estimates very precisely.
Lack of precision can stem from a number of factors, most obviously
a poor model fit to historical data,
low model sensitivity to one or more parameters, or
correlations between parameter estimates.

@@

We choose the resampling approach to estimating parameter precision
instead of the more-common technique
of estimating confidence intervals from standard errors for two reasons.
First, model parameters are highly constrained, and confidence intervals often violate the constraints.
For example, a result of $\alpha_1 = 0.25 \pm 0.3$ is nonsensical,
because $ \alpha_1 \in [0, 1]$.
Second, true confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha_1 = 0.2 + 0.1, -0.05$),
but the standard error approach yields symmetric confidence intervals.
The resampling approach that we employed both respects constraints
and allows for asymmetric confidence intervals.

@@
